k-plus Anticlustering: An Improved k-means Criterion for Maximizing Between-Group Similarity

### 2024-03-06 paper.Rmd
some on abstract, probably does not make it better

    Numerous applications in psychological research require that a data set is partitioned via the maximization---instead of the more common minimization---of a clustering criterion. Such anticlustering seeks for high similarity between groups (i.e., maximum diversity) and/or high pairwise dissimilarity within groups (i.e., maximum dispersion). Brusco et al. (2020) proposed a bicriterion algorithm (BILS) that simultaneously seeks for maximum diversity and dispersion, introducing the bicriterion approach for anticlustering. Here, we advance the bicriterion approach in several ways. As two minor extensions, we allow that the dispersion and the diversity can be optimized on the basis of different dissimilarity matrices, and that other anticlustering objectives of between-group similarity can be optimized instead of only the diversity. As our main contribution, we present an exact algorithm that returns multiple globally optimal partitions with regard to the dispersion. Despite its theoretical computational hardness, our method scales to rather large data sets (e.g., $N > 5000$ for $K = 3$ groups). By extending existing optimal and heuristic approaches (including BILS), we evaluate how to best obtain high diversity on top of an optimal dispersion. Our analysis reveals an important factor determining the difficulty of the bicriterion anticlustering problem: the level of restriction induced by the maximum dispersion constraint. Depending on the level of restriction, different algorithmic choices are recommended. The original BILS as well as our extensions are available via the free and open source R package `anticlust`. Practical examples are included to illustrate their application.

>>>>

    Numerous applications in psychological research require that a data set is partitioned via the maximization---instead of the more common minimization---of a clustering criterion. Such anticlustering seeks for high similarity between groups (i.e., maximum diversity) and/or high pairwise dissimilarity within groups (i.e., maximum dispersion). Brusco et al. (2020) proposed a bicriterion algorithm (BILS) that simultaneously seeks for maximum diversity and dispersion, introducing the bicriterion approach for anticlustering. Here, we advance the bicriterion approach in several ways. As a heuristic, the BILS does not necessarily identify globally optimal criterion values. We combine a novel exact algorithm for the dispersion with the BILS heuristic to obtain high diversity on top of an optimal dispersion. We present several adaptations of the BILS and evaluate which is best suited to accomplish this goal. Despite its theoretical computational hardness, our optimal method scales to large data sets (e.g., $N > 5000$ for $K = 3$ groups). Our analysis reveals an important factor determining the difficulty of the bicriterion anticlustering problem: the level of restriction induced by the maximum dispersion constraint. Depending on the level of restriction, different algorithmic choices are recommended. In particular, the more constrained the problem becomes, the more important is the iterated local search (ILS), the second phase of the BILS. The original BILS as well as our extensions are available via the free and open source R package `anticlust`. Practical examples are included to illustrate their application.
